### 2011-06-08 thesis/novelty.tex
Add how to use a MAP to perform novelty detection between 2 graphs.
This starts introducing the relation between graph-structure and novel probability.

\section{Detecting Novelty on a Single Variable}
This section focus on detecting a novel class on single variable of the
conceptual map. As seen before on \autoref{chap:novelty-intro},
novelty detection can be done by implementing a threshold on
$P(x|\overline{novel})/P(x)$.
In order to correctly measure novelty on a single variable both
conditional and unconditional probabilities need to be coherently defined:

@@

Let $U_a$ be this new graph based on $G$ with variable $a$ replaced by $a^*$.
$P_{U_a}(x)$ models then the unconditional distribution of the feature set $x$
without the assumption that $a$ is one of the known classes and so is used as
an approximation for $P(x)$.

@@

Having now defined $P_G$ and $P_{U_a}$ as models models for
$P(x|\overline{novel})$ and $P(x)$, the ratio $P_G(x)/P_{U_a}(x)$ is used to
define an order-relation for the novelty of $a$.
Now all that is missing is a threshold $t_A$ to trigger on.
Such a threshold can be obtained by calculating which value of $t_A$
triggers a desired amount of the labelled data.
Also, since there is no reason to distinguish between different instances of
a variable type, only one threshold per variable type needs to be trained.

Concluding: novelty on a variable $a$ of type $A$ should be detected when:

\begin{equation}
\frac{P_G(x)}{P_{U_a}(x)} < t_A
\end{equation}

where $t_A$ is a threshold for variables of type $A$, $P_G(x)$ denotes the
probability of the sensed variables $x$ being generated by a distribution
where $a$ is known, and $P_{Ua}$ denotes the probability of $x$ being generated
by a distribution where only $a$ is not necessarily a class the system has
learned before.

@@

Let the known states of $a$ be $\{A_1,\dotsc, A_{n}\}$ and the unconditional variable
$a^*$ be modelled with states $\{A^*_1,\dotsc, A^*_{k}\}$. I.e. if $k=1$ then the
unconditional variable is being modelled with an assumption on features that do
not dependent on the class.

First, extend variable $a$ with new states $A_{n+k}$

By incorporating novelty information back into the graphical model the presented
method allows

\section{Copping with Multiple Novel Variables}
On the previous section it was shown how to detect a novel class on a variable
assuming that the rest of the graph variables and structure is correctly modeled
by the agent knowledge.
In this section methods are presented that not only allow to extended this to
detection on multiple variables but also allow the system to incorporate
novelty information back in the graphical model in order to continue its
probabilistic and uncertain reasoning.

>>>>

\section{Detecting Novelty on a Single Variable}
This section focus on detecting a novel class on single variable of the
conceptual map. As seen before on \autoref{chap:novelty-intro},
assuming that $P(\overline{novel})$ is constant, novelty detection can be
done by implementing a threshold on $P(x|\overline{novel})/P(x)$.
In order to correctly measure novelty on a single variable both
conditional and unconditional probabilities need to be coherently defined:

@@

Let $U_a$ be this new graph based on $G$ with variable $a$ replaced by $a^*$.
$P_{U_a}(x)$ models then the unconditional distribution of the feature set $x$
without the assumption that $a$ is one of the known classes and can be is used
as an approximation for $P(x)$.

@@

Having now defined $P_G$ and $P_{U_a}$ as models models for
$P(x|\overline{novel})$ and $P(x)$, the ratio $P_G(x)/P_{U_a}(x)$ is used to
define an order-relation for the novelty of $a$.
All that is missing is a threshold $t_A$ to trigger on.
Such a threshold can be obtained by calculating which value of $t_A$
triggers a desired amount of the labelled data.
Also, since there is no reason to distinguish between different instances of
a variable type, only one threshold per variable type needs to be trained.

Concluding: assuming a constant $P(\overline{novel})$ for any specific set and
structure of $x$, novelty on a variable $a$ of type $A$ should be detected when:

\begin{equation}
\label{eq:threshold-with-ta}
\frac{P_G(x)}{P_{U_a}(x)} < t_A
\end{equation}

where $t_A$ is a threshold for variables of type $A$, $P_G(x)$ denotes the
probability of the sensed variables $x$ being generated by a distribution
where $a$ is known, and $P_{Ua}$ denotes the probability of $x$ being generated
by a distribution where $a$ is not necessarily a class the system has
learned before.

@@

The method presented before obtains an unconditional model
from the conditional model by replacing a variable $a \in \{A_1,\dotsc, A_n\}$
and the associated factors $\phi_a$ with a variable
$a^* \in \{A^*_1,\dots,A^*_k\}$ and related unconditional factors $\phi_{a^*}$
and leaving all the remaining variables and factors equal.

With this it becomes possible to merge both graphs into a single one $G$ by
merging $a$ and $a^*$ into $b$ and factors $\phi_a$ and $\phi_{a^*}$ into
$\phi_b$ as follows:

\[
b \in \{A_1,\dotsc,A_n,A^*_1,\dotsc,A^*_k\}
\]

\[
\phi_b(b, \cdot) = \left\{
  \begin{array}{l l}
    \phi_a(b, \cdot), & \quad \text{if } b \in \{A_1,\dotsc,A_n\} \\
    \phi_{a^*}(b, \cdot), & \quad \text{if } b \in \{A^*_1,\dotsc,A^*_k\}
  \end{array} \right.
\]

Additionally it is possible to control variable $b$ states into being only one
of the states of $a$ or $a^*$ by introducing a control variable $c \in {0, 1}$
and factor $\phi_{c}$:
\[
\phi_{c}(b, c) = \left\{
  \begin{array}{l l}
    1, & \qquad \text{if } c = 0 \text{ and } b \in \{A_1,\dotsc,A_n\} \\
    1, & \qquad \text{if } c = 1 \text{ and } b \in \{A^*_1,\dotsc,A^*_n\}\\
    0, & \qquad \text{otherwise}
  \end{array} \right.
\]

By conditioning the control variable $c$ it is possible to obtain a graph with
potentials on any of the remaining variables equivalent to the potentials
obtained with $G_A$ (i.e.\ $c=0$) or $G_{U_a}$ (i.e.\ $c=1$). By introducing
an extra threshold factor $\phi_t(c)$ it also becomes possible to balance the
potential of the distribution described by $B$ such that a \gls{MAP} operation
on the $c$ variable decides on the novelty of $a$ with a desired threshold:

\begin{eqnarray*}
 P_B(c=0 | x)         &<& P_B(c=1 | x) \\
\Phi_B(c=0 \land x)   &<& \Phi_B(x=1 \land x) \\
\Phi_G(x) \phi_t(c=0) &<& \Phi_{U_a}(x) \phi_t(c=1) \\
P_G(x) Z_G \phi_t(c=0) &<& P_{U_a}(x) Z_{U_a} \phi_t(c=1) \\
\frac{P_G(x)}{P_{U_a}(x)} &<& \frac{\phi_t(c=1)}{\phi_t(c=0)}\frac{Z_{U_a}}{Z_G}
\end{eqnarray*}

Concluding, after merging graph $G$ and $U_a$ into graph $B$, the condition
described in \autoref{eq:threshold-with-ta} can be test by performing a
\gls{MAP} operation on variable $c$ of graph $B$ when the threshold factor
$\phi_t$ is proportional to:
\begin{equation}
\phi_t(c) \propto \left\{
  \begin{array}{l l}
    1, & \qquad \text{if } c = 0 \\
    t_A \frac{Z_G}{Z_{U_a}}, & \qquad \text{if } c = 1\\
  \end{array} \right.
\end{equation}

This equation is not very practical due to the need of calculation of
normalization factors $Z_G$ and $Z_{U_a}$. The presence of this factors
not only turns computation more expensive, but also requires that a full model
is known for all the factors disabling the ability to incorporate probabilistic
sensing output from complex detectors such as \glspl{SVM}. Ultimately, it also
removes capability to mix complex thresholds on several variables that would
allow to do probabilistic inference on the novelty of multiple variables at once.

\begin{figure}[h]
\centering
lalla
\caption{\label{fig:merging-graphs}A condition based on the ratio between 2
         factor graphs that share some structure can be performed by testing the
         MAP probability of a variable on a merged graph.}
\end{figure}

Analysing the reason for the normalization factors playing a role on the
threshold can be tracked down to the assumption that $P(novel)$ is
considered constant over any set or structure of $x$.

This requirement forces the threshold function to use the normalization factors
to compensate for a bigger likelihood of certain structure being naturally more
disposed to generate novel categories.

This relation between graph structure and $P(novel)$ is discussed in more detail
in the next section where also the novelty detection method is made to work
without normalization factors, allowing probabilistic sensing, and extended to
detect novelty on multiple variables.

\section{Detecting and Copping with Novelty on Multiple Variable}
On the previous section it was shown how to detect a novel class on a variable
assuming that the rest of the graph variables and structure is correctly modeled
by the agent knowledge.
In this section methods are presented that not only allow to extended this to
detection on multiple variables but also allow the system to incorporate
novelty information back in the graphical model in order to continue its
probabilistic and uncertain reasoning.
